### 2022-04-05 paper/localpsf.tex
Massive cuts and edits to intro

@@

Since our end goal is an H-matrix, it is natural to ask why we do not build the H-matrix directly using classical H-matrix construction methods. The reason is that \emph{classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free}. We overcome this difficulty using impulse response interpolation. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

@@

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to inverse problems, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
%In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. 
More accessible Hessian approximations are highly desirable because they permit fast solution of the minimization problem via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for the parameter \cite{LAPLACEAPPROXIMATION}.

@@

The most popular matrix-free operator approximation methods are based on low rank approximation \cite{CCGOPAPERS}. This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. However, low rank approximation is not suitable for high rank operators, and often operators of practical interest are high rank. In the context of Hessians, methods based on low rank approximation suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator actions are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}.

Product-convolution (PC) approximations are approximations of an operator by weighted sums of convolution operators with spatially varying weights. A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. PC approximations have been used in a variety of fields going back several decades \cite{PRODCONVLIST}. Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope. However, ``PSF'' is sometimes used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features. For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. A common choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. However, computing each impulse response required one action of $\Aop$. Even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, were required.

In this paper, we reduce the computational cost further by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. This idea was inspired by techniques from resolution analysis in seismic imaging \cite{RESOLUTION}. In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, for instance, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION}, the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

Conventionally, PC approximations are applied to vectors using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. One may perform these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. Thus, it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. 
While not commonly done, in theory linear algebra operations involving convolution-product approximations (transposes of PC approximations) could be done using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET} since these are equivalent to low rank approximation of the pseudo-differential symbol of the operator. We also note other relevant matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} which do not use PC approximation.

@@

In this paper, we build on previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format. Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to make the following improvements:
\begin{itemize}
	\item PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while the method we propose is based on an assumption that we call ``local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). 
	\item PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. 
\end{itemize}
In exchange for these improvements, the proposed approximation cannot be written as a weighted sum of convolution operators, hence we cannot apply this approximation to vectors via the FFT. However, since our approximation is an H-matrix, it can be applied to vectors efficiently via H-matrix methods.

Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, which we do not use here. These alternative methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

>>>>

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
%In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. 
%More accessible Hessian approximations are highly desirable because they permit fast solution of the minimization problem via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for the parameter \cite{LAPLACEAPPROXIMATION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{CCGOPAPERS}. 
%This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator actions are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}.

@@

%Since our end goal is an H-matrix, it is natural to ask why we do not build the H-matrix directly using classical H-matrix construction methods. The reason is that 
Classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free. Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS,ILONAHMATRIX,NOEMIHMATRIX}, which we do not use here. 
%These alternative methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. 
While methods based on the peeling process are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

We overcome this difficulty using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate shifted versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

Picking the batches of points $x_i$ requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{RESOLUTION}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

%We pick batches of points $x_i$ in such that many $\phi_{x_i}$ are computed with each operator action of $\Aop$.

Our impulse response interpolation procedure may be categorized as a ``point spread function'' (PSF) method that is loosely based on so-called ``product-convolution'' (PC) approximations, which are are approximations of an operator by weighted sums of convolution operators with spatially varying weights \cite{PRODCONVLIST}. PC and PSF methods have a long history dating back several decades; for background we recommend the following papers: \cite{PRODCONVGOOD}. While PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, the method we propose is based on an assumption that we call ``local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). Furthermore, PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined.

%A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. 
%The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. 
%Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope, and the local convolution kernels are shifted\footnote{The impulse responses are shifted so that each convolution kernel is centered at zero instead of the location of the point source.} impulse responses of the system to point sources of light.
%, which characterize the local distortion and blurring, and the weighting functions interpolate between these convolution kernels. 
%PSF is also used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features. For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

%The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses $\phi_{x_i}$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. 
%A common choice for PC methods based on scattered impulse responses is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation.
%In our previous work, we started with a coarse grid of points $x_i$, then adaptively refined the grid in the regions where the error in the approximation was large \cite{PRODCONVMYPAPER}. However, computing each $\phi_{x_i}$ required one action of $\Aop$. Even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, were required. In this paper, we reduce the computational cost further by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which
%In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, for instance, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION},
%the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{RESOLUTION}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

@@

@@

%PC approximations have been used in a variety of fields going back several decades \cite{PRODCONVLIST}; for background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}. The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. A common choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. However, computing each impulse response required one action of $\Aop$. Even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, were required.

% A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. PC approximations have been used in a variety of fields going back several decades \cite{PRODCONVLIST}. Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope. However, ``PSF'' is sometimes used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features. For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

%The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. A common choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. However, computing each impulse response required one action of $\Aop$. Even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, were required.

%In this paper, we reduce the computational cost further by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. This idea was inspired by techniques from resolution analysis in seismic imaging \cite{RESOLUTION}. In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, for instance, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION}, the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

%Conventionally, PC approximations are applied to vectors using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. One may perform these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. Thus, it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. 
%While not commonly done, in theory linear algebra operations involving convolution-product approximations (transposes of PC approximations) could be done using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET} since these are equivalent to low rank approximation of the pseudo-differential symbol of the operator. We also note other relevant matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} which do not use PC approximation.

@@

%In this paper, we build on previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format.

%In exchange for these improvements, the proposed approximation cannot be written as a weighted sum of convolution operators, hence we cannot apply this approximation to vectors via the FFT. However, since our approximation is an H-matrix, it can be applied to vectors efficiently via H-matrix methods.
